---
title: "STAT 184 Final Project"
names: "Mackenzie Johnson & Isabella Santori"
output: html_notebook
---

```{r}
# clean up the RStudio environment 
rm(list = ls())

# load all packages
library(mosaic)
library(tidyverse)
library(lubridate)
library(DataComputing)
```

# Research Question: Does being from a certian country make you more likely to medal at the Olypmics for Track and Field?

### Load Primary Data Source into RNotebook
```{r}
#Primary Data Soucre
olympicAthleteData <- read.csv(file.choose())
glimpse(olympicAthleteData)
```
This data set contains information about every athlete that has competed in the Olympics. A case is an Olympic athlete and there are 15 variables to describe each athlete and their participation in the Olympics. 

### Load Secondary Data Source into RNotebook
```{r}
library(rvest)
page <- "https://en.wikipedia.org/wiki/List_of_Olympic_records_in_athletics"
tableList <- page %>%
  read_html() %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)

glimpse(tableList[[1]])
glimpse(tableList[[2]])
```
These lists are from Wikipedia and they are a list of both the male and female current olympic records in the track and field events. Each case represents a different event and then has the current record and the athlete who currently hold that record. The tablelist 1 is the male records and the tablelist 2 is the female records. 

# Data Wrangling 
### Regular Expression of Olympic Records table 
```{r}
malerecords<- tableList[[1]]
femalerecords<- tableList[[2]]

malerecords <- 
  malerecords %>%
    filter(grepl(pattern = "\\d*metres$", Event)) %>%
    mutate(Record = ms(Record)) %>%
    mutate(Sex = "M")
malerecords

femalerecords <-
  femalerecords %>%
    filter(grepl(pattern = "\\d*metres$", Event)) %>%
    mutate(Record = ms(Record)) %>%
    mutate(Sex = "F")
femalerecords
```

### Joining Two Data Tables
```{r}
SprintingRecords <-
  malerecords %>%
    full_join(femalerecords)
SprintingRecords
```

### Using Data Verbs on primary data set
```{r}
olympicAthleteData <-
  olympicAthleteData%>%
  select(Name, Sex, Age, Team, Year, Sport, Event, Medal)%>%
  filter(Sport == "Athletics") %>%
  mutate(Name = as.character(Name),
         Sex = as.character(Sex),
         Sport = as.character(Sport),
         Event = as.character(Event),
         Medal = as.character(Medal),
         Team = as.character(Team)) %>%
  filter(grepl(pattern = "\\d*metres$", Event)) %>%
  mutate(Event = (gsub(pattern = "Athletics Women's|Athletics Men's", replacement = "", Event))) %>%
  rename("Athlete(s)" = Name) %>%
  rename("Nation" = Team)

olympicAthleteData
```

### Using Data verbs on secondary data set
```{r}
SprintingRecords <-
  SprintingRecords %>%
  select(Event, Record, Nation, Date, `Athlete(s)`, Sex) %>%
  mutate(Date = mdy(Date),
         Year = year(Date)) %>%
  select(Event, Record, Nation, Year, `Athlete(s)`, Sex) %>%
  mutate(Nation = (gsub(pattern = "\\(.{3}\\)$", replacement = "", Nation))) %>%
  mutate(Medal = "Gold")
  
SprintingRecords
```
### Fix names in secondary data set to match with primary data set
```{r}
SprintingRecords <-
SprintingRecords %>%
  mutate(`Athlete(s)` = gsub(pattern = "Usain Bolt", replacement = "Usain St. Leo Bolt", `Athlete(s)`)) %>%
  mutate(`Athlete(s)` = gsub(pattern = "David Rudisha", replacement = "David Lekuta Rudisha", `Athlete(s)`)) %>%
  mutate(`Athlete(s)` = gsub(pattern = "Noah Ngeny", replacement = "Noah Kiprono Ngeny", `Athlete(s)`)) %>%
  mutate(`Athlete(s)` = gsub(pattern = "Kenenisa Bekele", replacement = "Kenenisa Bekele Beyecha", `Athlete(s)`)) %>%
  mutate(`Athlete(s)` = gsub(pattern = "Florence Griffith Joyner", replacement = "Delorez Florence Griffith Joyner", `Athlete(s)`)) %>%
  mutate(`Athlete(s)` = gsub(pattern = "Marie-José Pérec", replacement = "Marie Jos- Juliana Prec \\(Hontas-\\)", `Athlete(s)`)) %>%
  mutate(`Athlete(s)` = gsub(pattern = "Nadezhda Olizarenko", replacement = "Nadiya Fedorivna Olizarenko \\(Mushta-\\)", `Athlete(s)`)) %>%
  mutate(`Athlete(s)` = gsub(pattern = "Paula Ivan", replacement = "Paula Ivan \\(Ilie\\)", `Athlete(s)`)) %>%
  mutate(`Athlete(s)` = gsub(pattern = "Vivian Cheruiyot", replacement = "Vivian Jepkemoi Cheruiyot", `Athlete(s)`)) %>%
  mutate(`Athlete(s)` = gsub(pattern = "Almaz Ayana", replacement = "Almaz Ayana Eba", `Athlete(s)`))

SprintingRecords
```

### Join primary and secondary data set
```{r}
completeOlympicData<- 
  olympicAthleteData%>%
  full_join(SprintingRecords)%>%
  arrange(Record) %>%
  select(`Athlete(s)`,Sex, Nation, Event, Record, Year, Medal)

completeOlympicData
```

### Making the data glyph ready 
```{r}
AthleteCounts <-
completeOlympicData %>%
  group_by(Nation, Sex) %>%
  summarise(count = n()) %>%
  spread(key = Sex, value = count) %>%
  mutate(TotalAthletes = F + M) %>%
  arrange(desc(TotalAthletes)) %>%
  na.omit()
AthleteCounts

MedalCounts <-
  completeOlympicData %>%
    group_by(Nation, Medal) %>%
    summarise(count = n()) %>%
    spread (key = Medal, value = count) %>%
    mutate(TotalMedals = sum(Bronze + Gold + Silver)) %>%
    select(Bronze, Gold, Silver, TotalMedals) %>% 
    arrange(desc(TotalMedals)) %>%
    na.omit()
MedalCounts

Counts <-
  MedalCounts %>%
    left_join(AthleteCounts)
Counts
```
# Graphics

### Plot Athlete Counts vs. Medal Counts by Country
```{r}
Counts %>%
  ggplot(aes(x = TotalAthletes, y = TotalMedals)) + 
  geom_point(aes(color = Nation)) + 
  theme(legend.position = "right") + 
  labs(title = "Athlete Counts vs. Medal Counts") 
```
This scatterplot demonstrates that countries with more total track and field athletes tend to earn more total medals. 

### Bar Graph of Medal Counts by Country and Event

